---
title: "RLumModel - Using own parameter sets"
author: "Johannes Friedrich"
date: '`r Sys.Date()`'
output:
  pdf_document:
    fig_height: 5
    fig_width: 5
    number_sections: yes
header-includes: \usepackage[utf8]{inputenc}
bibliography: /home/johannes/RLumModel.bib
csl: /home/johannes/elsevier-harvard.csl
toc: true
vignette: |
  %\VignetteIndexEntry{RLumModel - Using own parameter sets} 
  %\VignetteEncoding{UTF-8}{inputenc}
  %\VignetteEngine{knitr::rmarkdown} 
---
```{r, echo=FALSE, message = FALSE}
rm(list = ls(all=TRUE))
library(RLumModel)
```

# Introduction

With RLumModel $\geq$ 0.2.0 it is possible to simulate quartz luminescence behaviour of own parameter sets or parameter sets, which are not included in the package but also known in literature. 
This little manual gives two examples how to implement parameter sets and prooves the results simulating the original simulations.

RLumModel offers maximum flexibility and fast calculation of first-order ordinary differential equations (ODEs), because of:

* flexible handling of different numbers of electron traps and hole centres in generating the model equations. So the user has not to care about coding ODEs in R. 
* Solving these equations is performed with C++ [@Eddelbuettel_2013] and offers a fast calculation of simulation results.

# Examples

This chapter shows the handling of own parameter sets with RLumModel. For this purpose three quartz luminescence models known from literature were taken, @Lawless_2009, @Pagonis_2009a and @Adamiec_2005. 



## Pagonis 2009 

@Pagonis_2009a presented parameters for their luminescence modelling of radioluminescence. This model was built for $Al_2O_3:C$ but the rate equations are identical. Below is a step-by-step manual for invlving these parameters in RLumModel and re-calculate the simulation made by @Pagonis_2009a. Note that in the original publication Figure 3 and Figure 6 are inconsistent with one another. For a doserate of 0.1 Gy/s an initial RL intensity of ca. 1.5e14 is obtained (see Figure 6 in original publication and simulations below). 


### Set parameters

First of all the model parameters had to be set. In RLumModel this can be done via list().
The list has to contain the following items:

* N: Concentration of electron- and hole traps [$cm^{-3}$]
* E: Electron/Hole trap depth [eV]
* s: Frequency factor [$s^(-1)$]
* A: Conduction band to electron trap and valence band to hole trap transition probability [$s^{-1} \cdot cm^3$].
* B: Conduction band to hole centre transition probability [$s^{-1} \cdot cm^3$].
* Th: Photo-eviction constant or photoionisation cross section, respectively
* E_th: Thermal assistence energy [eV]
* k_B: Boltzman constant 8.617e-05 [eV/K]
* W: activation energy 0.64 [eV] (for UV)
* K: 2.8e7 (dimensionless constant). 
* model: "customized"
* R (optional): Ionisation rate (pair production rate) equivalent to 1 Gy/s [$s^{-1} \cdot cm^{-3}$]

**Note**: 

* Not every publication uses the same definition of parameter A and B.
* When no thermal quenching is expected, set $\mathrm{K} = 0$. A numerical value of W is necessary, if K $\neq$ 0.
* The parameter model = "customized" is necessary not to load a stored model within the RLumModel pacakge.
* The luminescence center, which is responsible for the luminescence production (TL-, OSL-, RF-signal), has to be the last entry in the model parameters, see examples below.
* For further details of the parameter see @Bailey_2001 and @Wintle_1975.
* The first two entries in N, A and B belong to the electron traps and the last two entries to the hole centres. This is not necessary but recommended.
* The first entry in N, E, s, A, B, Th and E_th belong to the first energy level (e.g. 110°C), the second, third, ... entries to second, third, ... energy levels.

```{r}
own_parameters <- list(
  N = c(2e15, 2e15, 2.4e16, 1e17),
  E = c(0, 0, 0, 0),
  s = c(0, 0, 0, 0),
  A = c(2e-8, 2e-9, 4e-9, 1e-8),
  B = c(0, 0, 5e-11, 4e-8),
  Th = c(0, 0),
  E_th = c(0, 0),
  k_B = 8.617e-5,
  K = 0,
  model = "customized",
  R = 1.7e15)
```

It is important to notive, that in @Pagonis_2009a *B* is the valence band to hole centre probability, but in @Bailey_2001 this is $\mathrm{A_j}$. The default setting of RLumModel is the definition by @Bailey_2001 and so the values of *B* (in @Pagonis_2009a) are *A* in the notation above.

As a next stept it is possible to set own 'starting-parameters', also called state parameters. In the case of @Pagonis_2009a they submitted initial concentrations of electrons and holes. This can be done via:
  
```{r}  
own_state_parameters <- c(0, 0, 0, 9.4e15, 0, 0)
```

Here the first entry is the first electron trap, the second entry the second electron trap, the third entry the hole center
and the fourth entry the luminescence center responsible for the RF signal.
The vector "own_state_parameters" needs six entries to be compatible with RLumModel (2 electron traps, 2 hole traps, 1 valence band, 1 conduction band).
This is necessary for the used R-package deSolve [@Soetaert_2012], because it requires as much state conditions as rate equations.
  
### Running the simulation with RLumModel

When all parameters are set the simulation can be started. The maim function in RLumModel is 'model_LuminescenceSignals()' and the usage with own parameter sets are described below. For a general overview for creating a sequence, running RLumModel with stored models etc. the user is referred to @Friedrich_2016 and to the vignette 'RLumModel - Getting started with RLumModel'. For simulating the results of @Pagonis_2009a the follwing sequence is needed.

```{r set sequence Pagonis 2009}  
sequence <- list(RF = c(20, 0.1, 0.1))
```
  
This sequence describes a radiofluorescence simulation at 20°C with a dose of 0.1 Gy and a dose rate of 0.1 Gy/s, so the stimulation time is 1s. 
Now all parts are prepared to simulate this sequence. 
The parameters 'own_parameters' and 'own_state_parameters' in 'model_LuminescenceSignals()' are prepared for using own created parameter sets. As parameter 'model =  "customized"' is necessary to not load stored parameters. 
  
  
```{r, fig.align="center"}  
RF_Pagonis2009 <- model_LuminescenceSignals(
  model = "customized", 
  sequence = sequence, 
  own_parameters = own_parameters, 
  own_state_parameters = own_state_parameters,
  verbose = FALSE)
```
  
As in the original publication initially the RF signal increases and is followed by an approximately linear region until the stimulation ends.
Figure 5 in @Pagonis_2009a shows the concentration of the luminescence center m1 for the stimulation time of 1s.
With RLumModel this can be plotted very fast with the following command (for a detailed description see vignette "RLumModel - Getting started with RLumModel").
The ylim command in *Luminescence::plot_RLum* was used to generate the same plot as Figure 5 in @Pagonis_2009a.
  
```{r, echo=FALSE, fig.align="center"}
concentration_m1 <- Luminescence::get_RLum(
  RF_Pagonis2009, 
  recordType = c("conc. level 4"))
  
Luminescence::plot_RLum(
  concentration_m1, 
  ylim = c(9.2e15, 9.6e15))
```


### Re-calculate the original results

Reproducing Figure 3 and Figure 6 in @Pagonis_2009a a loop over different dose rates is necessary. 
The following code lines are able to run the model for five different dose rates from 0.1 to 0.5 Gy/s and plot all contained RF curves and the initial RF signal.
For a more detailed descripton of the loop and the single commands therein the user is referred to @Friedrich_2016 and the vignette 'RLumModel - Getting started with RLumModel'.

```{r Different Dose rates Pagonis 2009, fig.align="center"}
dose.rate <- seq(from = 0.1, to = 0.5, by = 0.1)

model.output <- lapply(dose.rate, function(x) {
    
    sequence <- list(RF = c(20, x, x))
    
    RF_data <- model_LuminescenceSignals(
      model = "customized", 
      sequence = sequence, 
      own_parameters = own_parameters, 
      own_state_parameters = own_state_parameters,
      verbose = FALSE,
      plot = FALSE
    )
    
    ## "RF$" for exact matching RF and not (RF)
    return(get_RLum(RF_data, recordType = "RF$", drop = FALSE))
    
  })

model.output.merged <- merge_RLum(model.output)

plot_RLum(
 object = model.output.merged,
 xlab = "Stimulation time [s]",
 ylab = "RF signal [a.u.]",
 legend.text = paste(dose.rate, "Gy/s"),
 legend.pos = "outside",
 combine = TRUE)
```



The following code calcultes the initial RF signal for the five different dose rates.


```{r Initial signal, fig.align="center"}
dose.rate <- seq(from = 0.1, to = 0.5, by = 0.1)

model.output <- vapply(X = dose.rate, FUN = function(x) {
    
    sequence <- list(RF = c(20, x, x))
    
    temp <- model_LuminescenceSignals(
      model = "customized", 
      sequence = sequence, 
      own_parameters = own_parameters, 
      own_state_parameters = own_state_parameters,
      verbose = FALSE,
      plot = FALSE
    )
    
    ## "RF$" for exact matching RF and not (RF)
    RF_curve <- get_RLum(temp, recordType = "RF$")
    
    return(max(get_RLum(RF_curve)[2,2]))
    
  }, FUN.VALUE = 1)

plot(
  dose.rate,
  model.output,
  type = "b",
  xlab = "Stimulation Time [s]",
  ylab = "Initial RF intensitiy [a.u.]"
  )
```


The results show that RLumModel is able to simulate the same results as published in @Pagonis_2009a with only little effort. All these examples can be modified to own needs, e.g. own sequences or own parameters.

## Lawless 2009

### Set parameters

```{r Lawless 2009 set parameters, fig.align="center"}
own_parameters <- list(
  N = c(1e14, 1e15),
  E = c(0, 0),
  s = c(0, 0),
  A = c(1e-13, 1e-14),
  B = c(0, 1e-7),
  Th = c(0, 0),
  E_th = c(0, 0),
  k_B = 8.617e-5,
  K = 0,
  model = "customized",
  R = 1e8)

sequence <- list(RF = c(20, 100, 1))

RF_Lawless_2009 <- model_LuminescenceSignals(
  model = "customized", 
  sequence = sequence, 
  own_parameters = own_parameters,
  verbose = FALSE,
  plot = FALSE)

concentration_n <- Luminescence::get_RLum(
  RF_Lawless_2009, 
    recordType = c("conc. level 1"))
```

```{r, echo=FALSE}
Luminescence::plot_RLum(
  concentration_n,
  ylim = c(0,15e8), lwd = 3)

t <- seq(0, 100, 2)
numerical_eq16 <- 1e-13*1e14/1e-7 *((1 + 2*1e-7*1e8*t/(1e-13*1e14))^(0.5)-1)

numerical_eq18 <- (2*1e-13*1e14*1e8*t/(1e-7))^(0.5)

lines(t, numerical_eq16, pch = 3, col = "red", type = "b")
lines(t, numerical_eq18, pch = 4, col = "green", type = "b")

legend("bottomright", legend = c("Simulated", "Eq. 16","Eq. 18"), col = c("black", "red", "green"), lwd = 1)
```



## Adamiec 2005


@Adamiec_2005 published model parameters for simulating thermal activation characteristics (TACs). 



### Set parameters

```{r Adamiec 2005 set parameters}
own_parameters <- list(
  N = c(1e8, 1e9, 5e10, 1.4e11, 1.8e11, 2e11, 1e11),
  E = c(0.7, 1.6, 4.0, 1.16, 1.75, 2.6, 3.2),
  s = c(4.7e8, 8.5e12, 0, 1.2e10, 7e14, 2e14, 0),
  A = c(2e-9, 3e-10, 2e-11, 4e-10, 6e-10, 1e-10, 1e-9),
  B = c(0, 0, 0, 3e-10, 3e-10, 2e-9, 1e-9),
  Th = c(0, 0, 0),
  E_th = c(0, 0, 0),
  k_B = 8.617e-5,
  K = 0,
  model = "customized",
  R = 1e9)
```


### Calculate TACs

```{r Adamiec 2005 TAC, echo=FALSE, fig.align="center"}

act.temp <- seq(from = 200, to = 700, by = 20)

##loop over temperature
model.output <- vapply(X = act.temp, FUN = function(x) {
  
sequence <- list(IRR = c(20, 1, 1), 
                 PH = c(x, 1000, 5), 
                 IRR = c(20, 0.01, 0.01), 
                 TL = c(20, 180, 2))
  
TAC_Adamiec2005 <- model_LuminescenceSignals(
  model = "customized", 
  sequence = sequence, 
  own_parameters = own_parameters, 
  verbose = FALSE,
  plot = FALSE)


  ## "TL$" for exact matching TL and not (TL)
  TL_curve <- get_RLum(TAC_Adamiec2005, recordType = "TL$")

  ##return max value in TL curve
  return(max(get_RLum(TL_curve)[,2]))

}, FUN.VALUE = 1)


plot(
  act.temp,
  model.output,
  type = "b",
  xlab = "Temperature [\u00B0C]",
  ylab = "TL [a.u.]"
  )
```



# Summary

# References